`Total-evidence' Bayesian estimation of phylogeny, divergence times and fossil ages

### 2016-04-05 response-to-reviewers.tex

Referee reports:

\section*{Referee: 1}

Comments to Author(s)
This paper is timely, as there has been a recent surge in interest in modelling morphological evolution to date species divergence times. The topic is also very interesting. However, many papers have obtained very ancient divergence time estimates under the method. For example, Beck and Lee (2014, PRSB, 281:20141278) found very ancient estimates of mammal divergence times, and estimates had also very large uncertainties. Other works report similar patters (e.g. Arcila et al. 2014, MPE, 82:131). People are very puzzled about whether these ancient date estimates are reliable, or whether the Mk model (which is simply the Jukes-Cantor extended to k states) is too simplistic. I therefore was a bit surprised when I read the abstract of the present ms, as it portrays the method very optimistically. However, after having examined the ms, I can see that levels of uncertainty in estimated fossil ages are very large (e.g. Figure 7). I therefore think the optimistic tone needs to be moderated.

\response{We have changed the abstract to moderate the tone. Specifically we have added the sentence ``We find that the estimation accuracy of fossil ages is generally high with credible intervals seldom excluding the true age, although the credible intervals were often wide, reflecting low precision.''}

@@

Overall I find the paper interesting. One strength is the modelling of the fossil ages as distributions based on stratigraphy, an important point we have recently highlighted (O'Reilly et al. 2015, TIGs, in press). I also find interesting the comparison between the stratigraphic age of fossils and the Bayesian estimates using the birth-death process.

I think the paper can be improved in certain areas. A bit of text needs to be added to clarify certain things (perhaps the ms is too short). Below I make several suggestions.

p.1. Abstract: change ``phylogenetic age'' by ``phylogenetic estimates of fossil age''. Please change throughout the ms.

\response{We have replaced all instances of ``phylogenetic age'' with variants of ``phylogenetic estimates of fossil age''.}

@@

$$p(tip\_ages) p(node\_ages) L(D|ages,tree)$$

where $p(tip\_ages)$ is the prior density on fossil ages using uniform distributions, $p(node\_ages)$ is the prior on internal node ages given the BD process, and $L(D|ages,tree)$ is the likelihood of the morphological data D, given the ages and the topology calculated using Lewis (2001) formula. However, it seems from the text in p.5 that another likelihood term may be involved. For the stepwise deletion procedure, I am guessing the priors are something like $p(tip\_ages\hat{}) p(node\_ages, tip\_age*)$, where now $tip\_ages\hat{}$ includes all the fossil ages except the one under analysis, which has been moved into the other BD-based prior term $(tip\_age*)$. Please provide general forms for the posteriors and make the connection with equation (1).

Give some more info on how the prior on the root age is set. Is Tmax the same as the root age? If it is, please state explicitly.

Please give the prior on the morphological substitution rate, and for relaxed-clock models, the prior on the rate variation parameter (I am assuming you are using log-normal clock, so prior on log-variance).

@@

p.8 paragraph starting on line 43. Do you mean that with the particular implementation of marginal likelihood calculation in BEAST you cannot calculate the BF? Or is this a general problem with any implementation of stepping-stones and this model? I am also confused, do sampling times refer to the ages of fossil tips? Aren't these unknown parameters being estimated in the MCMC? Then they would appear as terms in the prior and the likelihood. (?)

p.9 Results section. For both data sets the authors state that the relaxed clock model fits the data better, but I can't find the BF for this. Also, for both data sets, the authors have sections titled "Penguins conform well to a strict morphological clcok" and "Canids conform well to a morphological clock", but these headings seem to contradict the statement in lines 48-49 p.12 of discussion. I am also puzzled as the authors have not quantified here departure from the clock.

Please provide a table with prior and posterior mean estimates of the age of the root for both data sets under all models used, and 95\% CI. Similarly, in the same table provide prior and posterior estimates of mean morphological substitution rate for all data sets and models (relaxed and strict clock). For the relaxed-clocks, provide prior and posterior estimates of rate-variation parameter. Then, when the authors discuss whether the data fit or not the strict clock, they should refer to the estimates in the table.

@@

Figure 3 and 6: Mention the dataset in the legend (canids? penguins?).

Figure 6, panels (b), (c) and (d) seem empty.

Figure 7: Add dots indicating posterior means.

Figure 8: Please replace this with the consensus tree and provide 95\% CIs of internal node ages.

@@

I could not find references to Figures 3 and 8 in the text.

In the results section, please provide some quantification of the uncertainty in time estimates. For example,  the authors could provide estimates of the precision as (95\% CI width)/(posterior mean of fossil age) and express this as a percentage. Roughly, looking at figure 7, it seems that for some nodes the 95\% CI limits would range from anywhere between 50\% to 150\% of the point stratigraphic estimate. Clearly, uncertainty here is very, very large.

### 2016-04-05 fossilDating-revision.tex
removed instances of phylogenetic age

% 250 words
\abstract{
Recent advances have allowed for both morphological fossil evidence and molecular sequences to be integrated into a single combined inference of divergence dates under the rule of Bayesian probability. 
In particular the fossilized birth-death tree prior and the Lewis-MK model of evolution of discrete morphological change allow for the estimation of both divergence times and phylogenetic relationships between fossil and extant taxa. 
We exploit this statistical framework to investigate the internal consistency of these models by estimating the phylogenetic age of each fossil in turn, within two rich and well-characterized data sets of fossil and extant species.
We find that the estimation accuracy of fossil ages is generally high with credible intervals seldom excluding the true age, although the credible intervals were often wide, reflecting low precision. 
In fact in the two data sets we analyze the {\em phylogenetic age} of a fossil species is on average  $< 2$ My from the midpoint age of the geological strata from which it was excavated. 
The high level of internal consistency found in our analyses suggests that the Bayesian statistical model employed is a good fit for both the geological and morphological data, and provides evidence from real data that the framework used can accurately model the evolution of discrete morphological traits coded from fossil and extant taxa. 
We anticipate that this approach will have diverse applications beyond divergence time dating, including dating fossils that are temporally unconstrained, testing the ``morphological clock'', 
and for uncovering potential model misspecification and/or data errors when controversial phylogenetic hypotheses are obtained based on combined divergence dating analyses.
}

@@

It may be less widely appreciated by molecular evolutionary biologists that the statistical phylogenetic revolution in molecular evolution has also been mirrored in the increasing application of statistical phylogenetic reasoning in macroevolutionary and systematic studies of the fossil record \cite{Foote1996,huelsenbeck1997maximum, tavare2002using, WagnerMarcot2013}. 
%This includes applying phylogenetic reasoning to questions of biogeography, character evolution, and the completeness of the fossil record.
Here we extend this tradition of applying phylogenetic reasoning to the fossil record by focusing on the question of what phylogenetic inference techniques can tell us about the ``phylogenetic age'' of a fossil, based solely on its morphological characteristics and through them, its phylogenetic and temporal relationships with a set of reference fossils.

@@

We extend this approach into the realm of discrete morphological evolution by presenting a statistical model of evolution that generates an expectation on the distribution of fossils, their morphological characters. 
This model has been previously presented in the context of divergence time dating \cite{gavr2014,gavryushkina2015bayesian}. 
It is distinct from alternative divergence time dating approaches in that it provides an explicit treatment of the temporal information contained in fossil remains, whether or not related molecular sequence data is available. 
This leads to an estimate of the age of the most recent common ancestor of a group of fossil and extant taxa. A key difference between this approach and earlier approaches to tip-calibrated ``total-evidence'' dating \cite{Ronq2012} is the admission of a probability that each fossil taxon may represent a sampled ancestor of one or more taxa  in the tree \cite{gavr2014}.
We exploit this framework to attempt the estimation of the phylogenetic age of individual fossils based solely on morphological data and their phylogenetic affinities to related taxa of known age. 
The method is applied to two rich and well-characterized morphological data sets: (i) 
%TS how many extant penguins?
19 extant penguins and 36 fossil relatives \cite{ksepka2010,ksepka2012}, (ii) a sample of nine extant canids and 123 fossil relatives \cite{Slater2015}.

@@

Gavryushkina {\it et al} \cite{gavryushkina2015bayesian} described a ``total-evidence'' approach \cite[implemented in BEAST2;][]{Beast2} for phylogenetic estimation of time-trees that employs both fossil and molecular sequence data as equal partners under the rule of probability. 
We extend their work further by investigating the consistency between the estimated phylogenetic age of a fossil and the corresponding fossil age range determined by geological and biostratigraphic evidence. 
The model of time-tree phylogeny employed is the so-called fossilized birth-death process \cite{Heath2014}, which forms a prior probability distribution on the space of sampled-ancestor trees \cite{Gavr2013}.

A previous study described how a set of fossils with discrete morphological characters could be used to estimate a time-tree. 
Here we additionally allow for one or more of the fossils to have broad uninformative priors on their age. 
This allows for the age of some of the fossils to be estimated solely based on their morphological characters and the phylogenetic affinities of their morphology to other fossils with known ages in the time-tree. 
We refer to this estimate of a fossil's age as its {\em phylogenetic age}. 
In estimating each of the fossils' phylogenetic ages in turn, two questions can be answered: (i) How much information about an individual fossil's age is available from phylogenetic analysis of morphological data alone, and (ii) What is the level of phylogenetic evidence in support of the palaeontological age range for a fossil.
These two questions are investigated using two morphological data sets, one of 36 fossil penguins and their extant relatives \cite{ksepka2010,ksepka2012,gavryushkina2015bayesian}, and one of 123 canid fossils and their extant relatives \cite{Slater2015}.

\subsection*{Estimating the phylogenetic ages of penguin fossils}

@@

\subsection*{Estimating the phylogenetic ages of Canid fossils}

@@

We applied the M1 model to this data set assuming a fixed geological age for each taxon. Unlike the original study \cite{Slater2015} we did not apply any other constraints or priors on ancestral divergence times beyond the ages of the included fossils. 
As with the penguin data set we also performed an analysis for each of the 123 canid fossils in turn. Since we can't calculate Bayes factors without a fossil age range (see following section) we simply used the same priors on macroevolutionary parameters as employed by \cite{gavryushkina2015bayesian} in their \Mstrict{} model except with an upper limit on the origin of $T_\text{max} = 80$. To estimate the phylogenetic age of the focal fossil we sampled its age under the fossilized birth-death prior in the range $(0,T_\text{max})$ Mya.
%TS: which priors on d, r, s, T did we use now?

@@

\subsection*{Penguins conform well to a strict morphological clock}
Although \Mstrict{} is a very simple model, the phylogenetic age estimates for the penguin fossils were remarkably consistent with their palaeontological age ranges. 
Figure \ref{fig:phyloAgeVsGeoAge}a plots the geological age and range against the phylogenetic age estimates. The points in this plot have $R^2 = 0.903$. 
The median error (difference between the phylogenetic median and the geological median) is 1.96 Myr. A summary of the individual estimates are tabulated in Table \ref{fossilTable1}.

As judged by Bayes factors, only one fossil exhibited strong evidence (i.e. $\text{log BF} < -3.0$) that the phylogenetic age was inconsistent with the geological age range. The log BF for {\em Paraptenodytes antarcticus} was -3.4. In fact the majority of the fossils (23/36 = 64\%) had strong positive evidence for the geological age range (i.e. $\text{log BF} > 3.0$).
Likewise, if we consider only the posterior probability that the fossil is in the geological age range then three of the 36 fossils has a posterior probability $< 0.05$, suggesting low posterior support for the phylogenetic age being within the palaeontological age range.
These three fossils were {\em Madrynornis mirandus}, {\em Paraptenodytes antarcticus} and {\em Sphenicus muizoni} with posterior probabilities that the phylogenetic age is in the palaeontological range of 0.007, 0.001 and 0.001 respectively. 
All other fossils have posterior probabilities of $> 0.05$ of their age being in the palaeontological range. 
It is worth noting that the absolute discrepancy in the ages are still quite moderate for the three fossils with low posterior probabilities, with {\em M. mirandus}: 6.3Myr vs 10Myr (phylogenetic age versus palaeontological age), {\em P. antarcticus}: 29.9 vs 22, {\em S. muizoni}: 5.2 vs 9.1. 
The small posterior probabilities are partially caused in these cases because the corresponding palaeontological age range is narrow, apparently suggesting very precise geological knowledge of the ages of these three fossils.

@@

\Mrelaxed{} was the best-fitting model for the penguin data set according to the analysis of \cite{gavryushkina2015bayesian}. 
As with \Mstrict{} this model produced phylogenetic age estimates that were very concordant with the geological age ranges of the fossils (Figure \ref{fig:phyloAgeVsGeoAge}b), with an overall $R^2 = 0.924$. The median error was 2.05 Myr across all 36 fossils. 
In this analysis none of the fossils exhibited any evidence (i.e. $\text{log BF} < 0.0$) that the phylogenetic age was inconsistent with the geological age range. 
However if we consider the posterior probability that the fossil is in the geological age range then five of the 36 fossils had a posterior probability $< 0.05$ for \Mrelaxed{}, suggesting low posterior support for the phylogenetic age being within the palaeontological age range. 
These five fossils were {\em Madrynornis mirandus}, {\em Paraptenodytes antarcticus}, {\em Perudyptes devriesi}, {\em Sphenicus muizoni} and {\em Waimanu manneringi} with posterior probabilities that the phylogenetic age is in the palaeontological range of 0.035, 0.018, 0.046, 0.004, 0.037 respectively. 
All other fossils have posterior probabilities of $> 0.05$ of their age being in the palaeontological range. 
Again the absolute discrepancy in the ages are quite moderate for the five fossils with low posterior probabilities, with {\em M. mirandus}: 6.7Myr vs 10Myr (phylogenetic age versus palaeontological age), {\em P. antarcticus}: 28.0 vs 22, {\em P. devriesi}: 49.0 vs 40, {\em S. muizoni}: 5.1 vs 9.1 and {\em W. manneringi}: 56.7 vs 61.05. 
A summary of all the individual estimates are tabulated in Table \ref{fossilTable8}. 
The individual marginal posterior distributions of phylogenetic age under \Mrelaxed{} and the corresponding geological range are shown in Figures \ref{hist8_younger} and \ref{hist8_older}.

\begin{figure}
\includegraphics{Figure1.pdf}
\caption{\label{fig:phyloAgeVsGeoAge}
The Bayesian phylogenetic age estimate (median of marginal posterior) for each of the 36 penguin fossils plotted against their palaeontological age estimates, under two alternative site and molecular clock models. 
The palaeontological age estimates are represented by the mid-point of the range and the upper and lower limits. 
The Bayesian estimates are represented by the median of the marginal posterior distribution and the upper and lower limits of the 95\% HPD interval. 
The blue line shows the $x=y$. If the vertical line doesn't cross $x=y$, then the midpoint of the geological range is not in the phylogenetic 95\% HPD. 
If the horizontal line doesn't cross $x=y$, then the median phylogenetic estimate is not contained in the palaeontological age range. 
The three labelled fossils have posterior probability of less than 0.05 for their age being within the palaeontological age interval.
%TS do these three fossils not intersect in horizontal AND vertical?
}
\end{figure}

\begin{figure}
\includegraphics[width=5in]{Figure2.pdf}
\caption{A plot of the number of non-ambiguous morphological sites for the taxon against the precision of the phylogenetic age for (a) \Mstrict{} and (b) \Mrelaxed{} (i.e. the precision is 1/variance in the marginal posterior distribution of the age).}
\end{figure}

\begin{figure}
\includegraphics[width=6in]{8_penguins_eNprior/8_fossilDatingHist_younger.pdf}
\caption{\label{hist8_younger}Marginal posterior density plots for the phylogenetic age estimate of each of the 18 penguin fossils younger than 30 Myr using \Mrelaxed{}. Red boxes are the superimposed age ranges derived from geological data.}
\end{figure}

\begin{figure}
\includegraphics[width=6in]{8_penguins_eNprior/8_fossilDatingHist_older.pdf}
\caption{\label{hist8_older}Marginal posterior density plots for the phylogenetic age estimate of each of the 18 penguin fossils older than 30 Myr using \Mrelaxed{}. Red boxes are the superimposed age ranges derived from geological data.}
\end{figure}

@@

Overall the results of analyzing the penguin data set with the M1 and M8 models were strikingly concordant. 
Figure \ref{fig:compareM1M8} shows four regressions between the two models: (a) Regression of estimated phylogenetic age of \Mstrict{} against \Mrelaxed{}, (b) Regression of the error in the phylogenetic age estimate of \Mstrict{} against \Mrelaxed{} (c) Regression of posterior probability of palaeontological range of \Mstrict{} against \Mrelaxed{}, (d) Regression of Bayes factor (BF) for palaeontological range of \Mstrict{} against \Mrelaxed{}. 
%TS is the following quantified? otherwise I'd delete the next sentence!?
Under \Mrelaxed{} all fossils have positive evidence for their geological age range, whereas under \Mstrict{} there are a handful of fossils with negative evidence for the corresponding geological age range. Furthermore, assuming the median geological age is the truth, the variance in the phylogenetic estimation error of the fossil ages is larger under \Mstrict{} than under \Mrelaxed{}.
%AJD added
This evidence, along with the previous result that \Mrelaxed{} has a higher marginal likelihood than \Mstrict{} \cite{gavryushkina2015bayesian} suggests that the relaxed model is overall a better fit to the data.

\begin{figure}
\includegraphics{Figure_comparison.pdf}
\caption{\label{fig:compareM1M8}
Comparison of \Mstrict{} against \Mrelaxed{}. (a) Estimated phylogenetic age of \Mstrict{} against \Mrelaxed{} with $x=y$ line (blue), (b) Regression of error in estimated phylogenetic age of \Mstrict{} against \Mrelaxed{}, (c) Regression of posterior probability of palaeontological range of \Mstrict{} against \Mrelaxed{}, (d) Regression of Bayes factor (BF) for palaeontological range of \Mstrict{} against \Mrelaxed{}.}
\end{figure}

@@

The canid data set shows remarkably good consistency between palaeontological ages and phylogenetic ages, even with the simple strict morphological clock model (\Mstrict{}).
The $R^2 = 0.901$ between the phylogenetic and palaeontological ages (see Figure \ref{fig:canidMorphVsGeo}). 
Only 16 out of 123 fossils (13\%) don't have the paleontological age in the credible interval of the phylogenetic age and there are no extreme outliers. 
The median error is $1.66$My, which is even more accurate than the penguin data set.
This data set contains half as many morphological characters as does the penguin data set (122 versus $245$) but despite that the individual age estimates have more precision 
%TS why 99\% for candids and 95\% for penguins? 
%AJD I will check this later.
(median HPD range = 6.5 My for canids as opposed to 9.6My for penguins).

\begin{figure}
\includegraphics{canids/1c_phyloAgeVsGeoAge.pdf}
\caption{\label{fig:canidMorphVsGeo}
The Bayesian phylogenetic age estimate (median and credible interval of marginal posterior) for each of the 123 canid fossils plotted against their palaeontological age point estimates, under a strict morphological clock model \Mstrict{}. 
The Bayesian estimates are represented by the median of the marginal posterior distribution and the upper and lower limits of the 99\% HPD interval. Blue line shows the $x=y$. 
If the vertical line doesn't cross $x=y$, then the geological age is not in the credible interval of the phylogenetic age. The 16 fossils for which this is the case are labelled.}
%TS I find this plot very hard to read as many fossils are overlaying...can we do something there for better visualization?
\end{figure}

@@

The authors would like to thank Daniel Ksepka and Graeme T. Lloyd for insightful comments and suggestions on early drafts of this manuscript. AJD and TS were partially funded by Marsden grant UOA1324 from the Royal Society of New Zealand (\url{http://www.royalsociety.org.nz/programme?s/funds/marsden/awards/2013-awards/}). AJD was funded by a Rutherford Discovery Fellowship from the Royal Society of New Zealand (\url{http://www.royalsociety.org.nz}).  TS was supported in part by the European Research Council under the 7th Framework Programme of the European Commission (PhyPD; grant agreement 335529). AJD and TS also thank ETH Z\"{u}rich for funding.

>>>>

% 250 words
\abstract{
Recent advances have allowed for both morphological fossil evidence and molecular sequences to be integrated into a single combined inference of divergence dates under the rule of Bayesian probability. 
In particular the fossilized birth-death tree prior and the Lewis-MK model of evolution of discrete morphological change allow for the estimation of both divergence times and phylogenetic relationships between fossil and extant taxa. 
We exploit this statistical framework to investigate the internal consistency of these models by producing phylogenetic estimates of the age of each fossil in turn, within two rich and well-characterized data sets of fossil and extant species.
We find that the estimation accuracy of fossil ages is generally high with credible intervals seldom excluding the true age, although the credible intervals were often wide, reflecting low precision. 
In fact in the two data sets we analyze the phylogenetic estimates of fossil age is on average  $< 2$ My from the midpoint age of the geological strata from which it was excavated. 
The high level of internal consistency found in our analyses suggests that the Bayesian statistical model employed is a good fit for both the geological and morphological data, and provides evidence from real data that the framework used can accurately model the evolution of discrete morphological traits coded from fossil and extant taxa. 
We anticipate that this approach will have diverse applications beyond divergence time dating, including dating fossils that are temporally unconstrained, testing the ``morphological clock'', 
and for uncovering potential model misspecification and/or data errors when controversial phylogenetic hypotheses are obtained based on combined divergence dating analyses.
}

@@

It may be less widely appreciated by molecular evolutionary biologists that the statistical phylogenetic revolution in molecular evolution has also been mirrored in the increasing application of statistical phylogenetic reasoning in macroevolutionary and systematic studies of the fossil record \cite{Foote1996,huelsenbeck1997maximum, tavare2002using, WagnerMarcot2013}. 
%This includes applying phylogenetic reasoning to questions of biogeography, character evolution, and the completeness of the fossil record.
Here we extend this tradition of applying phylogenetic reasoning to the fossil record by focusing on the question of what phylogenetic inference techniques can tell us about the age of a fossil, based solely on its morphological characteristics and through them, its phylogenetic and temporal relationships with a set of reference fossils.

@@

We extend this approach into the realm of discrete morphological evolution by presenting a statistical model of evolution that generates an expectation on the distribution of fossils, their morphological characters. 
This model has been previously presented in the context of divergence time dating \cite{gavr2014,gavryushkina2015bayesian}. 
It is distinct from alternative divergence time dating approaches in that it provides an explicit treatment of the temporal information contained in fossil remains, whether or not related molecular sequence data is available. 
This leads to an estimate of the age of the most recent common ancestor of a group of fossil and extant taxa. A key difference between this approach and earlier approaches to tip-calibrated ``total-evidence'' dating \cite{Ronq2012} is the admission of a probability that each fossil taxon may represent a sampled ancestor of one or more taxa  in the tree \cite{gavr2014}.
We exploit this framework to attempt the estimation of the age of individual fossils based solely on morphological data and their phylogenetic affinities to related taxa of known age. 
The method is applied to two rich and well-characterized morphological data sets: (i) 
%TS how many extant penguins?
19 extant penguins and 36 fossil relatives \cite{ksepka2010,ksepka2012}, (ii) a sample of nine extant canids and 123 fossil relatives \cite{Slater2015}.

@@

Gavryushkina {\it et al} \cite{gavryushkina2015bayesian} described a ``total-evidence'' approach \cite[implemented in BEAST2;][]{Beast2} for phylogenetic estimation of time-trees that employs both fossil and molecular sequence data as equal partners under the rule of probability. 
We extend their work further by investigating the consistency between the phylogenetic estimate of the age of a fossil and the corresponding fossil age range determined by geological and biostratigraphic evidence. 
The model of time-tree phylogeny employed is the so-called fossilized birth-death process \cite{Heath2014}, which forms a prior probability distribution on the space of sampled-ancestor trees \cite{Gavr2013}.

A previous study described how a set of fossils with discrete morphological characters could be used to estimate a time-tree. 
Here we additionally allow for one or more of the fossils to have broad uninformative priors on their age. 
This allows for the age of some of the fossils to be estimated solely based on their morphological characters and the phylogenetic affinities of their morphology to other fossils with known ages in the time-tree. 
We refer to this as the phylogenetic estimate of the fossil's age. 
In phylogenetically estimating the age of each of the fossils in turn, two questions can be answered: (i) How much information about an individual fossil's age is available from phylogenetic analysis of morphological data alone, and (ii) What is the level of phylogenetic evidence in support of the palaeontological age range for a fossil.
These two questions are investigated using two morphological data sets, one of 36 fossil penguins and their extant relatives \cite{ksepka2010,ksepka2012,gavryushkina2015bayesian}, and one of 123 canid fossils and their extant relatives \cite{Slater2015}.

\subsection*{Phylogenetic estimates of the ages of penguin fossils}

@@

\subsection*{Phylogenetic estimates of the ages of Canids fossils}

@@

We applied the M1 model to this data set assuming a fixed geological age for each taxon. Unlike the original study \cite{Slater2015} we did not apply any other constraints or priors on ancestral divergence times beyond the ages of the included fossils. 
As with the penguin data set we also performed an analysis for each of the 123 canid fossils in turn. Since we can't calculate Bayes factors without a fossil age range (see following section) we simply used the same priors on macroevolutionary parameters as employed by \cite{gavryushkina2015bayesian} in their \Mstrict{} model except with an upper limit on the origin of $T_\text{max} = 80$. To phylogenetically estimate the age of the focal fossil we sampled its age under the fossilized birth-death prior in the range $(0,T_\text{max})$ Mya.
%TS: which priors on d, r, s, T did we use now?

@@

\subsection*{Penguins conform well to a strict morphological clock}
Although \Mstrict{} is a very simple model, the phylogenetic estimates of the ages of the penguin fossils were remarkably consistent with their palaeontological age ranges. 
Figure \ref{fig:phyloAgeVsGeoAge}a plots the geological age and range against the phylogenetic estimates of fossil age. The points in this plot have $R^2 = 0.903$. 
The median error (difference between the phylogenetic median and the geological median) is 1.96 Myr. A summary of the individual estimates are tabulated in Table \ref{fossilTable1}.

As judged by Bayes factors, only one fossil exhibited strong evidence (i.e. $\text{log BF} < -3.0$) that the phylogenetic estimate of fossil age was inconsistent with the geological age range. The log BF for {\em Paraptenodytes antarcticus} was -3.4. In fact the majority of the fossils (23/36 = 64\%) had strong positive evidence for the geological age range (i.e. $\text{log BF} > 3.0$).
Likewise, if we consider only the posterior probability that the fossil is in the geological age range then three of the 36 fossils has a posterior probability $< 0.05$, suggesting low posterior support for the phylogenetic estimate of fossil age being within the palaeontological age range.
These three fossils were {\em Madrynornis mirandus}, {\em Paraptenodytes antarcticus} and {\em Sphenicus muizoni} with posterior probabilities that the phylogenetic estimate of fossil age is in the palaeontological range of 0.007, 0.001 and 0.001 respectively. 
All other fossils have posterior probabilities of $> 0.05$ of their age being in the palaeontological range. 
It is worth noting that the absolute discrepancy in the ages are still quite moderate for the three fossils with low posterior probabilities, with {\em M. mirandus}: 6.3Myr vs 10Myr (phylogenetic estimate of fossil age versus palaeontological age), {\em P. antarcticus}: 29.9 vs 22, {\em S. muizoni}: 5.2 vs 9.1. 
The small posterior probabilities are partially caused in these cases because the corresponding palaeontological age range is narrow, apparently suggesting very precise geological knowledge of the ages of these three fossils.

@@

\Mrelaxed{} was the best-fitting model for the penguin data set according to the analysis of \cite{gavryushkina2015bayesian}. 
As with \Mstrict{} this model produced phylogenetic estimates of fossil age that were very concordant with the geological age ranges of the fossils (Figure \ref{fig:phyloAgeVsGeoAge}b), with an overall $R^2 = 0.924$. The median error was 2.05 Myr across all 36 fossils. 
In this analysis none of the fossils exhibited any evidence (i.e. $\text{log BF} < 0.0$) that the phylogenetic estimate of fossil age was inconsistent with the geological age range. 
However if we consider the posterior probability that the fossil is in the geological age range then five of the 36 fossils had a posterior probability $< 0.05$ for \Mrelaxed{}, suggesting low posterior support for the phylogenetic estimate of fossil age being within the palaeontological age range. 
These five fossils were {\em Madrynornis mirandus}, {\em Paraptenodytes antarcticus}, {\em Perudyptes devriesi}, {\em Sphenicus muizoni} and {\em Waimanu manneringi} with posterior probabilities that the phylogenetic estimate of fossil age is in the palaeontological range of 0.035, 0.018, 0.046, 0.004, 0.037 respectively. 
All other fossils have posterior probabilities of $> 0.05$ of their age being in the palaeontological range. 
Again the absolute discrepancy in the ages are quite moderate for the five fossils with low posterior probabilities, with {\em M. mirandus}: 6.7Myr vs 10Myr (phylogenetic estimate of fossil age versus palaeontological age), {\em P. antarcticus}: 28.0 vs 22, {\em P. devriesi}: 49.0 vs 40, {\em S. muizoni}: 5.1 vs 9.1 and {\em W. manneringi}: 56.7 vs 61.05. 
A summary of all the individual estimates are tabulated in Table \ref{fossilTable8}. 
The individual marginal posterior distributions of phylogenetic estimates of fossil age under \Mrelaxed{} and the corresponding geological range are shown in Figures \ref{hist8_younger} and \ref{hist8_older}.

\begin{figure}
\includegraphics{Figure1.pdf}
\caption{\label{fig:phyloAgeVsGeoAge}
The Bayesian phylogenetic estimate of fossil age (median of marginal posterior) for each of the 36 penguin fossils plotted against their palaeontological age estimates, under two alternative site and molecular clock models. 
The palaeontological age estimates are represented by the mid-point of the range and the upper and lower limits. 
The Bayesian estimates are represented by the median of the marginal posterior distribution and the upper and lower limits of the 95\% HPD interval. 
The blue line shows the $x=y$. If the vertical line doesn't cross $x=y$, then the midpoint of the geological range is not in the phylogenetic 95\% HPD. 
If the horizontal line doesn't cross $x=y$, then the median phylogenetic estimate is not contained in the palaeontological age range. 
The three labelled fossils have posterior probability of less than 0.05 for their age being within the palaeontological age interval.
%TS do these three fossils not intersect in horizontal AND vertical?
}
\end{figure}

\begin{figure}
\includegraphics[width=5in]{Figure2.pdf}
\caption{A plot of the number of non-ambiguous morphological sites for the taxon against the precision of the phylogenetic estimate of fossil age for (a) \Mstrict{} and (b) \Mrelaxed{} (i.e. the precision is 1/variance in the marginal posterior distribution of the age).}
\end{figure}

\begin{figure}
\includegraphics[width=6in]{8_penguins_eNprior/8_fossilDatingHist_younger.pdf}
\caption{\label{hist8_younger}Marginal posterior density plots for the phylogenetic estimate of fossil age of each of the 18 penguin fossils younger than 30 Myr using \Mrelaxed{}. Red boxes are the superimposed age ranges derived from geological data.}
\end{figure}

\begin{figure}
\includegraphics[width=6in]{8_penguins_eNprior/8_fossilDatingHist_older.pdf}
\caption{\label{hist8_older}Marginal posterior density plots for the phylogenetic estimate of fossil age of each of the 18 penguin fossils older than 30 Myr using \Mrelaxed{}. Red boxes are the superimposed age ranges derived from geological data.}
\end{figure}

@@

Overall the results of analyzing the penguin data set with the M1 and M8 models were strikingly concordant. 
Figure \ref{fig:compareM1M8} shows four regressions between the two models: (a) Regression of estimated phylogenetic estimates of fossil ages of \Mstrict{} against \Mrelaxed{}, (b) Regression of the error in the phylogenetic estimates of fossil ages of \Mstrict{} against \Mrelaxed{} (c) Regression of posterior probability of palaeontological range of \Mstrict{} against \Mrelaxed{}, (d) Regression of Bayes factor (BF) for palaeontological range of \Mstrict{} against \Mrelaxed{}. 
%TS is the following quantified? otherwise I'd delete the next sentence!?
Under \Mrelaxed{} all fossils have positive evidence for their geological age range, whereas under \Mstrict{} there are a handful of fossils with negative evidence for the corresponding geological age range. Furthermore, assuming the median geological age is the truth, the variance in the phylogenetic estimation error of the fossil ages is larger under \Mstrict{} than under \Mrelaxed{}.
%AJD added
This evidence, along with the previous result that \Mrelaxed{} has a higher marginal likelihood than \Mstrict{} \cite{gavryushkina2015bayesian} suggests that the relaxed model is overall a better fit to the data.

\begin{figure}
\includegraphics{Figure_comparison.pdf}
\caption{\label{fig:compareM1M8}
Comparison of \Mstrict{} against \Mrelaxed{}. (a) Phylogenetic estimate of fossil age of \Mstrict{} against \Mrelaxed{} with $x=y$ line (blue), (b) Regression of error in phylogenetic estimate of fossil age of \Mstrict{} against \Mrelaxed{}, (c) Regression of posterior probability of palaeontological range of \Mstrict{} against \Mrelaxed{}, (d) Regression of Bayes factor (BF) for palaeontological range of \Mstrict{} against \Mrelaxed{}.}
\end{figure}

@@

The canid data set shows remarkably good consistency between palaeontological ages and phylogenetic estimates of fossil ages, even with the simple strict morphological clock model (\Mstrict{}).
The $R^2 = 0.901$ between the phylogenetic and palaeontological ages (see Figure \ref{fig:canidMorphVsGeo}). 
Only 16 out of 123 fossils (13\%) don't have the paleontological age in the credible interval of the phylogenetic estimate of fossil age and there are no extreme outliers. 
The median error is $1.66$My, which is even more accurate than the penguin data set.
This data set contains half as many morphological characters as does the penguin data set (122 versus $245$) but despite that the individual age estimates have more precision 
%TS why 99\% for candids and 95\% for penguins? 
%AJD I will check this later.
(median HPD range = 6.5 My for canids as opposed to 9.6My for penguins).

\begin{figure}
\includegraphics{canids/1c_phyloAgeVsGeoAge.pdf}
\caption{\label{fig:canidMorphVsGeo}
The Bayesian phylogenetic estimate of fossil age (median and credible interval of marginal posterior) for each of the 123 canid fossils plotted against their palaeontological age point estimates, under a strict morphological clock model \Mstrict{}. 
The Bayesian estimates are represented by the median of the marginal posterior distribution and the upper and lower limits of the 99\% HPD interval. Blue line shows the $x=y$. 
If the vertical line doesn't cross $x=y$, then the geological age is not in the credible interval of the phylogenetic estimate of fossil age. The 16 fossils for which this is the case are labelled.}
%TS I find this plot very hard to read as many fossils are overlaying...can we do something there for better visualization?
\end{figure}

@@

The authors would like to thank Daniel Ksepka and Graeme T. Lloyd for insightful comments and suggestions on early drafts of this manuscript. We thank Mario do Reis and one anonymous reviewer for helpful comments during revision. AJD and TS were partially funded by Marsden grant UOA1324 from the Royal Society of New Zealand (\url{http://www.royalsociety.org.nz/programme?s/funds/marsden/awards/2013-awards/}). AJD was funded by a Rutherford Discovery Fellowship from the Royal Society of New Zealand (\url{http://www.royalsociety.org.nz}).  TS was supported in part by the European Research Council under the 7th Framework Programme of the European Commission (PhyPD; grant agreement 335529). AJD and TS also thank ETH Z\"{u}rich for funding.
